---
title: NYP Shootings
author: Fredy Paz
output:
  pdf_document: default
---

# NYPD Shootings Analysis

## Overview

The data set used in this analysis is from the City of New York. It contains detailed records of shooting incidents.

The intent of this document is to conduct analysis and gather insights. We will look at a breakdown of incidents per boroughs, date, and time of day.

Importing data into variable "nypddata" from <https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD>

```{r}
library(tidyverse)
nypddata <-
read.csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
```

Summary of data

```{r}
summary(nypddata)
```

## Data Cleaning

Removing columns PRECINT, JURISDICTION_CODE

```{r}
nypddatatidy <- nypddata %>% select(-JURISDICTION_CODE, -PRECINCT)
```

Changing OCCUR_DATE to Date format

```{r}
nypddatatidy <- nypddatatidy %>% mutate(OCCUR_DATE = mdy(OCCUR_DATE))
```

## Data Analysis

Group shootings by date, and provide a count of shooting by date

```{r}
nypd_bydate <- nypddatatidy %>% group_by(OCCUR_DATE) %>% summarize(total_incidents = n())
```

Adding a plot to view shootings by date

```{r}
ggplot(nypd_bydate, aes(x = OCCUR_DATE, y = total_incidents)) + geom_line(aes(color = "total_incidents")) + labs(title = "NYC Shootings by Date", x = "Date", y = "Total Incidents")
```

Getting exact date with most shootings

```{r}
maxdate <- nypd_bydate %>% filter(total_incidents == max(total_incidents))
```

Group shooting by Boroughs

```{r}
nypd_byboro <- nypddatatidy %>% group_by(BORO) %>% summarize(total_incidents = n())
```

Adding a bar chart to show shootings by Boroughs

```{r}

ggplot(nypd_byboro,aes(x = BORO, y = total_incidents)) + geom_col(fill = "red")
```

Adding a variable to change OCCUR_TIME to time format hms

```{r}
nypddatatidy <- nypddatatidy %>% mutate(time_hour = hour(hms(OCCUR_TIME)))
```

Grouping by hour of day and count shootings

```{r}
nypd_byhour <- nypddatatidy %>% group_by(time_hour) %>% summarise(total_incidents = n())
```

Adding a plot with linear model showing shootings by time of day

```{r}
ggplot(nypd_byhour, aes(x = time_hour, y = total_incidents)) +
    geom_point(color = "red") +  
    geom_smooth(formula = y ~ x, method = "lm", se = TRUE, color = "black", linetype = "dashed") +  
    labs(title = "Shootings by Time of Day with Linear Trend",
         x = "Hour of Day (24-hour format)",
         y = "Total Shootings")
```

The borough with the most shootings over the history of the data is Brooklyn with 11,346 total incidents.

```{r}
maxboro <- nypd_byboro %>% filter(total_incidents == max(total_incidents))
maxboro
```

The date with the most shootings was July 5th 2020 with 47 shootings

```{r}
maxdate
```

The bias that I identified was assuming that the most shootings occurred during peak covid times and that they would be mostly located in neighborhoods that we have historically associated with violence and poverty. I assumed Manhattan would not be at the top.

I reset my mindset to not make assumptions, relied on the data, and corroborated my findings by searching news for the specific date with most deaths. See resources below.

I found a few news articles mentioning the spike in violence but there was no single explanation or cause. It seems to be a combination of July 4th activity coupled with covid restrctions easing.

Resources:

<https://www.nbcnewyork.com/news/local/bullet-strikes-nypd-patrol-vehicle-misses-officers-sitting-inside/2500243/>

<https://abc7ny.com/nyc-shootings-2020-last-night-this-week-in/6299513/>
